---
title: "beat-side GAL4 in PNs"
output: html_notebook
---

This analysis was based on the GAL4 expression pattern data updated in May 2026. 

# Load packages. 
```{r}
library(tidyverse)
library(magrittr)
library(gplots)
library(RColorBrewer)
library(ggpubr)

library(pheatmap)
```


# Fig 3. Load the dataset and run the analyses based on the binary expression matrix. 
```{r}
df <- read_csv("PN beat-side expression matrix_2026_v2.csv") %>% as.data.frame() 
```

```{r}
df <- df[!is.na(df$Lineage), ]
```

```{r}
#For binary analysis in Fig 2, set all non-zero values to 1. For the analysis in Fig S8 based on the binned expression levels, keep the original matrix. 
df[, -(1:5)] <- (df[, -(1:5)] != 0) * 1
```

```{r}
mtx <- df[c("beat-Ia", 
            "beat-Ib",
            "beat-Ic",
            "beat-IIa",
            "beat-IIb",
            "beat-IIIa", 
            "beat-IIIb",
            "beat-IIIc",
            "beat-IV",
            "beat-Va", 
            "beat-Vc",
            "beat-VI", 
            "beat-VII",
            "side", 
            "side-II", 
            "side-III",
            "side-IV",
            "side-V",
            "side-VI",
            "side-VIII")] %>% as.matrix()

rownames(mtx) <- df[["Glomerulus"]]
```


## Fig 3G, H. Plot the adPN and lPN lineages seperately and order the PN class based on the birth sequence. 
```{r}
pnorder_ad <- c("DP1m", "VL2p", "DA4l", "VM4", "VA6", "VL2a", "DC1", "VA7l", "VA2", "DM4", "DL5", "VM3", "DM3", "DL4", "DL1", "DA3", "DC2", "D", "VA3", "DC3", "VA1d", "VM7v", "VM7d", "VM2", "DM6", "VA1v", "VC4", "VM5d", "VM5v", "VC3", "DL2v", "DL2d")

pnorder_l <- c("VA4", "VC2", "VC1", "DM1", "DM2", "VA5", "VA7m", "DA1", "DL3", "VM1", "DA2", "DM5")
```

```{r}
mtx_order_ad <- mtx[pnorder_ad, ]
mtx_order_l <- mtx[pnorder_l, ]
```


### Fig 3G, H. Expression heatmap. 
```{r, fig.width = 30, fig.height = 36}
#pdf("heatmap_revision_v2_PNorder_adPN.pdf", width = 30,height = 36)  

heatmap.2(mtx_order_ad, 
          Rowv = FALSE,
          Colv = TRUE,
          dendrogram = "none", 
          col = colorRampPalette(c("white", "dodgerblue"))(2),
          trace = "none",
          colsep = 0:ncol(mtx_order_ad),
          rowsep = 0:nrow(mtx_order_ad),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 3,  # Font size for column labels
          key = FALSE
          )

#legend("topleft", legend = unique(df$Lineage), fill = unique(row_colors), title = "Lineage", cex = 5)

#dev.off() 


#pdf("heatmap_revision_v2_PNorder_lPN.pdf", width = 30,height = 36)  

heatmap.2(mtx_order_l, 
          Rowv = FALSE,
          Colv = TRUE,
          dendrogram = "none", 
          col = colorRampPalette(c("white", "dodgerblue"))(2),
          trace = "none",
          colsep = 0:ncol(mtx_order_l),
          rowsep = 0:nrow(mtx_order_l),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 3,  # Font size for column labels
          key = FALSE
          )

#legend("topleft", legend = unique(df$Lineage), fill = unique(row_colors), title = "Lineage", cex = 5)

#dev.off()  
```


### Fig 3G, H. Correlation heatmap. 
```{r}
mtx_order_ad_cor <- cor(t(mtx_order_ad))

mtx_order_l_cor <- cor(t(mtx_order_l))
```

```{r, fig.asp=1}
#pdf("cormatrix_spectrum_v2_PNorder_adPN.pdf") 

pheatmap(mtx_order_ad_cor,
      color = colorRampPalette(c("skyblue2", "white", "tomato2"))(50), 
      cluster_rows = FALSE, 
      cluster_cols = FALSE, 
      show_rownames = TRUE, 
      show_colnames = TRUE, 
      fontsize = 14, 
      angle_col = 90, 
      border_color = "gray", 
      breaks = seq(-0.4, 1, length.out=51), 
      )

#dev.off() 


#pdf("cormatrix_spectrum_v2_PNorder_lPN.pdf") 

pheatmap(mtx_order_l_cor,
      color = colorRampPalette(c("skyblue2", "white", "tomato2"))(50), 
      cluster_rows = FALSE, 
      cluster_cols = FALSE, 
      show_rownames = TRUE, 
      show_colnames = TRUE, 
      fontsize = 14, 
      angle_col = 90, 
      border_color = "gray", 
      breaks = seq(-0.4, 1, length.out=51), 
      )

#dev.off() 
```


## Fig 3B. 
```{r, fig.width = 30, fig.height = 36}
row_colors <- brewer.pal(length(unique(df$Lineage)), "Set2")[as.numeric(as.factor(df$Lineage))]


#pdf("heatmap_spectrum_v2_OR.pdf", width = 30, height = 36)  

heatmap.2(mtx, 
          col = colorRampPalette(c("white", "dodgerblue"))(2),
          trace = "none",
          RowSideColors = row_colors,
          colsep = 0:ncol(mtx),
          rowsep = 0:nrow(mtx),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 3,  # Font size for column labels
          key = FALSE
          )

legend("topleft", legend = unique(df$Lineage), fill = unique(row_colors), title = "Lineage", cex = 5)

#dev.off()
```


## Fig 3C
```{r}
# Exclude the vPN glomerulus. 
df <- df[df$Lineage != "vPN", ]

mtx <- df[-c(1, 2, 3, 4, 5)] %>% as.matrix()

rownames(mtx) <- df[["Glomerulus"]]

mtx_rcor <- cor(t(mtx))
```


```{r}
similarity <- numeric()
relation <- character()

ndim = nrow(mtx_rcor)

for (i in 1:(ndim-1)) {
  for (j in (i+1):ndim) {
    similarity = append(similarity, mtx_rcor[i, j])
    
    if (df$Lineage[i] == df$Lineage[j]) {
        relation = append(relation, "within lineage")
    } else {
        relation = append(relation, "between lineages")
        }
  }
}
```


```{r}
similarity_summary <- data.frame(similarity = similarity, relation = relation)

similarity_summary$relation %<>% factor(levels = c("between lineages", "within lineage"))

#write_csv(similarity_summary, "Fig 3C_PN_similarity_summary.csv")
```


```{r}
similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between lineages", "within lineage"))
```

```{r}
#show text_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between lineages" = "#9BC985",
    "within lineage" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_v2.pdf")
```


## Fig 3D. 
```{r}
within_all <- numeric()
between_all <- numeric()

# Run multiple times of shuffling. 
for (x in 1776:2023) {
  
  set.seed(x)
  
  df_shuffled <- df[sample(1:nrow(df)), ]
  
  mtx <- df_shuffled[-c(1, 2, 3, 4, 5)] %>% as.matrix()
  
  mtx_rcor <- cor(t(mtx))
  
  within <- numeric()
  between <- numeric()
  
  ndim = nrow(mtx_rcor)
  
  for (i in 1:(ndim-1)) {
    
    for (j in (i+1):ndim) {
      
      if (df$Lineage[i] == df$Lineage[j]) {
        within = append(within, mtx_rcor[i, j])
       } else {
        between = append(between, mtx_rcor[i, j])}
    }
    
  }
  
  within_all = append(within_all, mean(within))
  between_all = append(between_all, mean(between))
  
}

# Summary of all shuffling runs.
similarity_summary_all <- data.frame(similarity = c(within_all, between_all), 
                                 relation = c(rep("within lineage", length(within_all)), 
                                              rep("between lineages", length(between_all))))

# Summary of the last one (with seed = 2023) as the representative shuffling result. 
similarity_summary <- data.frame(similarity = c(within, between), 
                                 relation = c(rep("within lineage", length(within)), 
                                              rep("between lineages", length(between))))

similarity_summary_all$relation %<>% factor(levels = c("between lineages", "within lineage"))
similarity_summary$relation %<>% factor(levels = c("between lineages", "within lineage"))

#write_csv(similarity_summary, "Fig 3D_PN_similarity_summary_shuffled.csv")
```


```{r}
similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary_all)


similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between lineages", "within lineage"))

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between lineages" = "#9BC985",
    "within lineage" = "#B595BF"
  )) + 
  #ylim(-0.5, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_v2_shuffle.pdf")
```




# Fig S8. Load the dataset and run the analyses based on the binned expression matrix. 
```{r}
df <- read_csv("PN beat-side expression matrix_2026_v2.csv") %>% as.data.frame()
```

```{r}
df <- df[!is.na(df$Lineage), ]
```

```{r}
mtx <- df[c("beat-Ia", 
            "beat-Ib",
            "beat-Ic",
            "beat-IIa",
            "beat-IIb",
            "beat-IIIa", 
            "beat-IIIb",
            "beat-IIIc",
            "beat-IV",
            "beat-Va", 
            "beat-Vc",
            "beat-VI", 
            "beat-VII",
            "side", 
            "side-II", 
            "side-III",
            "side-IV",
            "side-V",
            "side-VI",
            "side-VIII")] %>% as.matrix()

rownames(mtx) <- df[["Glomerulus"]]
```


## Fig S8D, E. 
```{r}
pnorder_ad <- c("DP1m", "VL2p", "DA4l", "VM4", "VA6", "VL2a", "DC1", "VA7l", "VA2", "DM4", "DL5", "VM3", "DM3", "DL4", "DL1", "DA3", "DC2", "D", "VA3", "DC3", "VA1d", "VM7v", "VM7d", "VM2", "DM6", "VA1v", "VC4", "VM5d", "VM5v", "VC3", "DL2v", "DL2d")

pnorder_l <- c("VA4", "VC2", "VC1", "DM1", "DM2", "VA5", "VA7m", "DA1", "DL3", "VM1", "DA2", "DM5")
```

```{r}
mtx_order_ad <- mtx[pnorder_ad, ]
mtx_order_l <- mtx[pnorder_l, ]
```

```{r}
mtx_order_ad_cor <- cor(t(mtx_order_ad))

mtx_order_l_cor <- cor(t(mtx_order_l))
```

```{r, fig.asp=1}
#pdf("cormatrix_spectrum_v2_PNorder_adPN.pdf") 

pheatmap(mtx_order_ad_cor,
      color = colorRampPalette(c("skyblue2", "white", "tomato2"))(50), 
      cluster_rows = FALSE, 
      cluster_cols = FALSE, 
      show_rownames = TRUE, 
      show_colnames = TRUE, 
      fontsize = 14, 
      angle_col = 90, 
      border_color = "gray", 
      #breaks = seq(-0.4, 1, length.out=51), 
      )

#dev.off() 


#pdf("cormatrix_spectrum_v2_PNorder_lPN.pdf") 

pheatmap(mtx_order_l_cor,
      color = colorRampPalette(c("skyblue2", "white", "tomato2"))(50), 
      cluster_rows = FALSE, 
      cluster_cols = FALSE, 
      show_rownames = TRUE, 
      show_colnames = TRUE, 
      fontsize = 14, 
      angle_col = 90, 
      border_color = "gray", 
      #breaks = seq(-0.4, 1, length.out=51), 
      )

#dev.off()
```


## Fig S8A. 
```{r, fig.width = 30, fig.height = 36}
row_colors <- brewer.pal(length(unique(df$Lineage)), "Set2")[as.numeric(as.factor(df$Lineage))]

pal <- brewer.pal(4, "Blues")
pal[1] <- "white"

#pdf("heatmap_spectrum_v2_OR.pdf", width = 30, height = 36)  

heatmap.2(mtx, 
          col = pal,
          trace = "none",
          RowSideColors = row_colors,
          colsep = 0:ncol(mtx),
          rowsep = 0:nrow(mtx),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 3,  # Font size for column labels
          key = FALSE
          )

legend("topleft", legend = unique(df$Lineage), fill = unique(row_colors), title = "Lineage", cex = 5)

#dev.off()
```


## Fig S8B. 
```{r}
# Exclude the vPN class.
df <- df[df$Lineage != "vPN", ]

mtx <- df[-c(1, 2, 3, 4, 5)] %>% as.matrix()

rownames(mtx) <- df[["Glomerulus"]]

mtx_rcor <- cor(t(mtx))
```


```{r}
similarity <- numeric()
relation <- character()

ndim = nrow(mtx_rcor)

for (i in 1:(ndim-1)) {
  for (j in (i+1):ndim) {
    similarity = append(similarity, mtx_rcor[i, j])
    
    if (df$Lineage[i] == df$Lineage[j]) {
        relation = append(relation, "within lineage")
    } else {
        relation = append(relation, "between lineages")
        }
  }
}
```


```{r}
similarity_summary <- data.frame(similarity = similarity, relation = relation)

similarity_summary$relation %<>% factor(levels = c("between lineages", "within lineage"))

#write_csv(similarity_summary, "Fig S8B_PN_similarity_summary_binned.csv")
```


```{r}
similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between lineages", "within lineage"))
```

```{r}
#show text_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between lineages" = "#9BC985",
    "within lineage" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_v2.pdf")
```


## Fig S8C
```{r}
within_all <- numeric()
between_all <- numeric()

# Run multiple times of shuffling. 
for (x in 1776:2023) {
  
  set.seed(x)
  
  df_shuffled <- df[sample(1:nrow(df)), ]
  
  mtx <- df_shuffled[-c(1, 2, 3, 4, 5)] %>% as.matrix()
  
  mtx_rcor <- cor(t(mtx))
  
  within <- numeric()
  between <- numeric()
  
  ndim = nrow(mtx_rcor)
  
  for (i in 1:(ndim-1)) {
    
    for (j in (i+1):ndim) {
      
      if (df$Lineage[i] == df$Lineage[j]) {
        within = append(within, mtx_rcor[i, j])
       } else {
        between = append(between, mtx_rcor[i, j])}
    }
    
  }
  
  within_all = append(within_all, mean(within))
  between_all = append(between_all, mean(between))
  
}

# Summary of all shuffling runs.
similarity_summary_all <- data.frame(similarity = c(within_all, between_all), 
                                 relation = c(rep("within lineage", length(within_all)), 
                                              rep("between lineages", length(between_all))))

# Summary of the last one (with seed = 2023) as the representative shuffling result. 
similarity_summary <- data.frame(similarity = c(within, between), 
                                 relation = c(rep("within lineage", length(within)), 
                                              rep("between lineages", length(between))))

similarity_summary_all$relation %<>% factor(levels = c("between lineages", "within lineage"))
similarity_summary$relation %<>% factor(levels = c("between lineages", "within lineage"))

#write_csv(similarity_summary, "Fig S8C_PN_similarity_summary_binned_shuffled.csv")
```


```{r}
similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary_all)


similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = TRUE), median(similarity, na.rm = TRUE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between lineages", "within lineage"))

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between lineages" = "#9BC985",
    "within lineage" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_v2_shuffle.pdf")
```