---
title: "beat-side GAL4 in ORNs"
output: html_notebook
---

This analysis was based on the GAL4 expression pattern data updated in April 2026. 

## Load packages. 
```{r}
library(tidyverse)
library(magrittr)
library(gplots)
library(RColorBrewer)
library(ggpubr)

library(showtext)
library(pheatmap)
```

## Load the dataset. 
```{r}
df <- read_csv("ORN beat-side expression matrix_2026.csv") %>% 
  as.data.frame() 

df$Sensillar_type <- factor(df$Sensillar_type, levels = c("TSB", "LB", "AT", "AC", "PB", "SA", "AR", "UNK"))
```

```{r}
#For binary analysis in Fig 2, set all non-zero values to 1. For the analysis in Fig S5 based on the binned expression levels, keep the original matrix. 

df_binary <- df

df_binary[, -(1:5)] <- (df_binary[, -(1:5)] != 0) * 1
```


## Fig 2G. Lineage-biased expression of beat/side genes.
```{r}
df_binary %>%
  group_by(Sensillar_type) %>%
  count()

expression_SensillarType <- df_binary %>%
  group_by(Sensillar_type) %>%
  summarise(across(where(is.numeric), mean)) 

expression_SensillarType <- expression_SensillarType[order(expression_SensillarType$Sensillar_type), ]
expression_SensillarType

#write_csv(expression_SensillarType, "expression_SensillarType.csv")
```


```{r}
expression_SensillarType_binarized <- expression_SensillarType %>% 
  mutate(across(where(is.numeric), ~ ifelse(. >= 0.6, 1, ifelse(. <= 0.4, 0, 0.5)))) 

mtx_template <- expression_SensillarType_binarized %>%
  select(-Sensillar_type) %>%
  as.matrix()

rownames(mtx_template) <- expression_SensillarType_binarized$Sensillar_type

pheatmap(mtx_template[(1:6), ],
      color = colorRampPalette(c("white", "black"))(3),
      cluster_rows = FALSE, 
      cluster_cols = FALSE, 
      show_rownames = TRUE, 
      show_colnames = TRUE, 
      fontsize = 14, 
      angle_col = 90, 
      #border_color = "gray", 
      )
```


```{r}
expression_SensillarType_binarized <- expression_SensillarType %>% 
  mutate(across(where(is.numeric), ~ ifelse(. >= 0.6, "present", ifelse(. <= 0.4, "absent", "either"))))
  #mutate(across(where(is.numeric), ~ ifelse(. > 0.5, "present", ifelse(. < 0.5, "absent", "either"))))

#write_csv(expression_SensillarType_binarized, "expression_SensillarType_binarized.csv")

expression_SensillarType_binarized_long <- expression_SensillarType_binarized %>%
  pivot_longer(cols = -Sensillar_type, names_to = "gene", values_to = "expression")


expression_SensillarType_binarized_long$gene <- 
  factor(expression_SensillarType_binarized_long$gene, 
         levels = c("beat-Ia", "beat-IIa", "beat-IIb", "side", "side-III", 
                    "beat-IIIb", "beat-IIIc", "beat-IV", "beat-VII", "side-VI",
                    "side-II", "side-V", "beat-Ib", 
                    "beat-Ic", "beat-Vc", "beat-VI", "side-IV", 
                    "beat-IIIa", "beat-Va", "side-VIII"))

expression_SensillarType_binarized_long %>%
  filter(Sensillar_type != "UNK" & Sensillar_type != "AR") %>%
  ggplot(aes(x = gene, y = Sensillar_type)) +
  scale_y_discrete(limits = rev) +
  geom_dotplot(aes(fill = expression), binaxis = "y", stackdir = "center", dotsize = 2) +
  #scale_fill_gradientn(colors = c("white", "gray", "black")) +
  scale_fill_manual(values = c("present"="black", "absent"="white", "either"="gray")) +
  geom_vline(xintercept = c(5.5, 12.5, 17.5), linewidth = 0.5) +
  theme(
    axis.title.x = element_blank(),     
    axis.title.y = element_blank(),    
    axis.text.x  = element_text(size = 14, angle = 45, vjust = 1, hjust = 1), 
    axis.text.y  = element_text(size = 14), 
    #legend.position = "none",
    legend.title = element_text(size = 12), 
    legend.text = element_text(size = 12), 
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.background = element_rect(fill = "white", color = NA)
  )

#ggsave("expression_template_split_revision1.pdf")
```


## Fig S12D. Expression breadth of beat/side genes based on the GAL4 labeling data across ORNs. 
```{r}
df_binary %>%
  summarise(across(where(is.numeric), sum)) %>%
  t() %>%
  as.data.frame()

df_binary %>%
  summarise(across(where(is.numeric), mean)) %>%
  t() %>%
  as.data.frame()


df_binary[df_binary$Sensillar_type %in% c("TSB", "AT", "LB", "AC"), ] %>%
  nrow()

df_binary[df_binary$Sensillar_type %in% c("TSB", "AT", "LB", "AC"), ] %>%
  summarise(across(where(is.numeric), sum)) %>%
  t() %>%
  as.data.frame()

df_binary[df_binary$Sensillar_type %in% c("TSB", "AT", "LB", "AC"), ] %>%
  summarise(across(where(is.numeric), mean)) %>%
  t() %>%
  as.data.frame()
```


## Fig 2B, 2C, 2D, S5. Hierarchical clustering analysis of beat/side expression across ORNs based on the binary/binned GAL4 labeling data. 

### Fig 2B.
```{r}
mtx <- df_binary[-c(1,2,3,4,5)] %>% as.matrix()

rownames(mtx) <- df_binary[["Glomerulus"]]
```


```{r, fig.width = 30, fig.height = 36}
row_colors <- brewer.pal(length(unique(df$Sensillar_type)), "Set2")[as.numeric(df$Sensillar_type)]

#pdf("heatmap_revision_jaccard.pdf", width = 30,height = 36)

heatmap.2(mtx, 
          col = colorRampPalette(c("white", "dodgerblue"))(2),
          dendrogram = "both",
          trace = "none",
          RowSideColors = row_colors,
          colsep = 0:ncol(mtx),
          rowsep = 0:nrow(mtx),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 4,  # Font size for column labels
          key = FALSE,
          #density.info = "none", 
          #keysize = 0.6,         # smaller than default (default ~1.5)
          #key.par = list(mar = c(3, 3, 1, 1))  # shrink margins around key
          )

legend("topleft", 
       legend = sort(unique(df$Sensillar_type)), 
       fill =  brewer.pal(length(unique(df$Sensillar_type)), "Set2"),
       title = "Sensillar types", 
       cex = 4.5)

#dev.off()  
```


### Fig 2C. 
```{r}
# Exclude two glomeruli that are innervated by more than one type of ORNs. 
df <- df_binary[df_binary$Glomerulus != "VP4" & df_binary$Glomerulus != "VL1", ] 

mtx <- df[-c(1,2,3,4,5)] %>% as.matrix()
rownames(mtx) <- df[["Glomerulus"]]

mtx_rcor <- cor(t(mtx))
```


```{r}
similarity <- numeric()
relation <- character()

ndim = nrow(mtx_rcor)

for (i in 1:(ndim-1)) {
  for (j in (i+1):ndim) {
    similarity = append(similarity, mtx_rcor[i, j])
    
    if (df$Sensillum[i] == df$Sensillum[j]) {
        relation = append(relation, "within sensillum")
    } else if (df$Sensillar_type[i] == df$Sensillar_type[j]) {
      relation = append(relation, "within sensillar type")
    } else {
        relation = append(relation, "between sensillar types")
        }
  }
}
```


```{r}
similarity_summary <- data.frame(similarity = similarity, relation = relation)

similarity_summary$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))

#write_csv(similarity_summary, "Fig 2C_ORN_similarity_summary.csv")
```


```{r}
similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between sensillar types", "within sensillar type"), 
                      c("between sensillar types", "within sensillum"), 
                      c("within sensillar type", "within sensillum"))
```


```{r}
showtext_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between sensillar types" = "#9BC985",
    "within sensillar type" = "#F7D58B",
    "within sensillum" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        )

#ggsave("similarity_comparison_spectrum.pdf")
```


### Fig 2D. 
```{r}
within_sensillum_all <- numeric()
within_type_all <- numeric()
between_type_all <- numeric()

# Run multiple times of shuffling. 
for (x in 1789:2024) {
  
  set.seed(x)
  df_shuffled <- df[sample(1:nrow(df)), ]
  
  mtx <- df_shuffled[-c(1,2,3,4,5)] %>% as.matrix()
  mtx_rcor <- cor(t(mtx))
  
  within_sensillum <- numeric()
  within_type <- numeric()
  between_type <- numeric()
  
  ndim = nrow(mtx_rcor)
  
  for (i in 1:(ndim-1)) {
    
    for (j in (i+1):ndim) {
      
      if (df$Sensillum[i] == df$Sensillum[j]) {
        within_sensillum = append(within_sensillum, mtx_rcor[i, j]) 
        } else if (df$Sensillar_type[i] == df$Sensillar_type[j]) { 
        within_type = append(within_type, mtx_rcor[i, j]) 
        } else {
        between_type <- append(between_type, mtx_rcor[i, j])
        }
    }
  }
 
  within_sensillum_all = append(within_sensillum_all, mean(within_sensillum))
  within_type_all = append(within_type_all, mean(within_type))
  between_type_all <- append(between_type_all, mean(between_type))
  }

# Summary of all shuffling runs.
similarity_summary_all <- data.frame(similarity = c(within_sensillum_all, within_type_all, between_type_all), 
                                 relation = c(rep("within sensillum", length(within_sensillum_all)), 
                                              rep("within sensillar type", length(within_type_all)), 
                                              rep("between sensillar types", length(between_type_all))
                                              )
                                 )

# Summary of the last one (with seed = 2024) as the representative shuffling result. 
similarity_summary <- data.frame(similarity = c(within_sensillum, within_type, between_type), 
                                 relation = c(rep("within sensillum", length(within_sensillum)), 
                                              rep("within sensillar type", length(within_type)), 
                                              rep("between sensillar types", length(between_type))
                                              )
                                 )

similarity_summary_all$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))
similarity_summary$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))

#write_csv(similarity_summary, "Fig 2D_ORN_similarity_summary_shuffled.csv")
```


```{r}
similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary_all)


similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between sensillar types", "within sensillar type"), 
                      c("between sensillar types", "within sensillum"), 
                      c("within sensillar type", "within sensillum"))


showtext_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between sensillar types" = "#9BC985",
    "within sensillar type" = "#F7D58B",
    "within sensillum" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_shuffle.pdf")
```




### Fig S5A. 
```{r}
# Reload the original binned dataset.
df <- read_csv("ORN beat-side expression matrix_2026.csv") %>% 
  as.data.frame() 

df$Sensillar_type <- factor(df$Sensillar_type, levels = c("TSB", "LB", "AT", "AC", "PB", "SA", "AR", "UNK"))

mtx <- df[-c(1,2,3,4,5)] %>% as.matrix()

rownames(mtx) <- df[["Glomerulus"]]
```


```{r, fig.width = 30, fig.height = 36}
row_colors <- brewer.pal(length(unique(df$Sensillar_type)), "Set2")[as.numeric(df$Sensillar_type)]


#pdf("heatmap_revision_jaccard.pdf", width = 30,height = 36)

pal <- brewer.pal(4, "Blues")
pal[1] <- "white"

heatmap.2(mtx, 
          col = pal,
          dendrogram = "both",
          trace = "none",
          RowSideColors = row_colors,
          colsep = 0:ncol(mtx),
          rowsep = 0:nrow(mtx),
          sepcolor = "gray", 
          sepwidth = c(0.01, 0.01),
          cexRow = 3,  # Font size for row labels
          cexCol = 4,  # Font size for column labels
          key = FALSE,
          #density.info = "none", 
          #keysize = 0.6,          # smaller than default (default ~1.5)
          #key.par = list(mar = c(3, 3, 1, 1))  # shrink margins around key
          )

legend("topleft", 
       legend = sort(unique(df$Sensillar_type)), 
       fill =  brewer.pal(length(unique(df$Sensillar_type)), "Set2"),
       title = "Sensillar types", 
       cex = 4.5)

#dev.off()  
```


### Fig S5B
```{r}
# Exclude two glomeruli that are innervated by more than one type of ORNs. 
df <- df[df$Glomerulus != "VP4" & df$Glomerulus != "VL1",]

mtx <- df[-c(1,2,3,4,5)] %>% as.matrix()
rownames(mtx) <- df[["Glomerulus"]]

mtx_rcor <- cor(t(mtx))
```


```{r}
similarity <- numeric()
relation <- character()

ndim = nrow(mtx_rcor)

for (i in 1:(ndim-1)) {
  for (j in (i+1):ndim) {
    similarity = append(similarity, mtx_rcor[i, j])
    
    if (df$Sensillum[i] == df$Sensillum[j]) {
        relation = append(relation, "within sensillum")
    } else if (df$Sensillar_type[i] == df$Sensillar_type[j]) {
      relation = append(relation, "within sensillar type")
    } else {
        relation = append(relation, "between sensillar types")
        }
  }
}
```


```{r}
similarity_summary <- data.frame(similarity = similarity, relation = relation)

similarity_summary$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))

#write_csv(similarity_summary, "Fig S5B_ORN_similarity_summary_binned.csv")
```


```{r}
similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between sensillar types", "within sensillar type"), 
                      c("between sensillar types", "within sensillum"), 
                      c("within sensillar type", "within sensillum"))
```


```{r}
showtext_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between sensillar types" = "#9BC985",
    "within sensillar type" = "#F7D58B",
    "within sensillum" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        )

#ggsave("similarity_comparison_spectrum.pdf")
```


### Fig S5C
```{r}
within_sensillum_all <- numeric()
within_type_all <- numeric()
between_type_all <- numeric()

# Run multiple times of shuffling. 
for (x in 1789:2024) {
  
  set.seed(x)
  df_shuffled <- df[sample(1:nrow(df)), ]
  
  mtx <- df_shuffled[-c(1,2,3,4,5)] %>% as.matrix()
  mtx_rcor <- cor(t(mtx))
  
  within_sensillum <- numeric()
  within_type <- numeric()
  between_type <- numeric()
  
  ndim = nrow(mtx_rcor)
  
  for (i in 1:(ndim-1)) {
    
    for (j in (i+1):ndim) {
      
      if (df$Sensillum[i] == df$Sensillum[j]) {
        within_sensillum = append(within_sensillum, mtx_rcor[i, j]) 
        } else if (df$Sensillar_type[i] == df$Sensillar_type[j]) { 
        within_type = append(within_type, mtx_rcor[i, j]) 
        } else {
        between_type <- append(between_type, mtx_rcor[i, j])
        }
    }
  }
 
  within_sensillum_all = append(within_sensillum_all, mean(within_sensillum))
  within_type_all = append(within_type_all, mean(within_type))
  between_type_all <- append(between_type_all, mean(between_type))
  }

# Summary of all shuffling runs.
similarity_summary_all <- data.frame(similarity = c(within_sensillum_all, within_type_all, between_type_all), 
                                 relation = c(rep("within sensillum", length(within_sensillum_all)), 
                                              rep("within sensillar type", length(within_type_all)), 
                                              rep("between sensillar types", length(between_type_all))
                                              )
                                 )

# Summary of the last one (with seed = 2024) as the representative shuffling result. 
similarity_summary <- data.frame(similarity = c(within_sensillum, within_type, between_type), 
                                 relation = c(rep("within sensillum", length(within_sensillum)), 
                                              rep("within sensillar type", length(within_type)), 
                                              rep("between sensillar types", length(between_type))
                                              )
                                 )

similarity_summary_all$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))
similarity_summary$relation %<>% factor(levels = c("between sensillar types", "within sensillar type", "within sensillum"))

#write_csv(similarity_summary, "Fig S5C_ORN_similarity_summary_binned_shuffled.csv")
```


```{r}
similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary_all %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary_all)


similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::count()

similarity_summary %>% 
  dplyr::group_by(relation) %>%
  dplyr::summarise(mean(similarity, na.rm = FALSE), median(similarity, na.rm = FALSE))

compare_means(similarity ~ relation, data = similarity_summary)
```


```{r}
my_comparison <- list(c("between sensillar types", "within sensillar type"), 
                      c("between sensillar types", "within sensillum"), 
                      c("within sensillar type", "within sensillum"))

showtext_auto()

data <- similarity_summary

ggplot(data, aes(x = relation, y = similarity, fill = relation)) +
  geom_boxplot() +
  #geom_violin() +
  #geom_jitter() +
  geom_signif(
    comparisons = my_comparison,
    na.rm = TRUE,
    #test = "t.test",
    step_increase = 0.1, 
    textsize = 4,
    map_signif_level = function(p) sprintf("%.2g", p),
    #y_position = 0.5
    ) +
  scale_fill_manual(values = c(
    "between sensillar types" = "#9BC985",
    "within sensillar type" = "#F7D58B",
    "within sensillum" = "#B595BF"
  )) + 
  #ylim(NA, 1) +
  #coord_fixed(ratio=1) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_text(size = 18),
        legend.title=element_text(size = 18),
        legend.text=element_text(size = 15)
        ) 

#ggsave("similarity_comparison_spectrum_shuffle.pdf")
```

